---
title: "Metrics-11-29-2020"
author: "Sixtus Dakurah"
date: "11/29/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
ctrl_dge_t <- read.table("/Users/sixtusdakurah/Desktop/Liger/data/pbmc_10X.txt")
stim_dge_t <- read.table("/Users/sixtusdakurah/Desktop/liger/data/pbmc_SeqWell.txt")
#ctrl_dge_t <- read_csv("pbmc_SqW.csv")
#stim_dge_t <- read_csv("pbmc_10X.csv")
head(ctrl_dge_t)
head(stim_dge_t)
```


## Triplet Loss

$$Loss = \sum_i^N \left[ ||f(c_i^a) - f(c_i^p)||_2^2 - ||f(c_i^a) - f(c_i^n)||_2^2 \right]_+$$
### Procedure

(1) Each cell $c_i$ is an anchor cell denoted as $c_i^a$.

(2) Select a hard positive $c_i^p$ - a cell in the same cluster as $c_i^a$, also select a hard negative $c_i^n$.

(3) Obtain the embeddings of the three cells from their right H matrices.

(4) Compute the loss.


Do this just for a sample of 10 cells from liger and neumf. 
(Make sure the anchor cells are the same in both cases - for the entire cells, this is not necessary).



```{r}
# load the cluster data
original_clusters <- read.csv("/Users/sixtusdakurah/Desktop/Liger/clusters.csv") #head(original_clusters)
# don't really need this
#liger_clusters <- read.csv("clusters-liger.csv") #head(liger_clusters)
#neumf_clusters <- read.csv("clusters-neumf.csv") #neumf clusters

# load the H-matrices from liger -- don't really need them
H1_Liger = read.csv("/Users/sixtusdakurah/Desktop/Liger/data/H1-Liger.csv")
H2_Liger = read.csv("/Users/sixtusdakurah/Desktop/Liger/data/H2-Liger.csv")

# load the H-matrices from neumf
H1_NeuMF = read.csv("/Users/sixtusdakurah/Desktop/Liger/data/H2-01-02-2021.csv") # NAMES ARE SWITCHED SOMEHOW
H2_NeuMF = read.csv("/Users/sixtusdakurah/Desktop/Liger/data/H1-01-02-2021.csv")
```


```{r}
head(original_clusters)
head(H1_Liger)
head(H2_Liger)
head(H1_NeuMF)
(H2_NeuMF)
```

```{r}
dim(H1_Liger)
dim(H2_Liger)
dim(H1_NeuMF)
dim(H2_NeuMF)
```


```{r}
# option 1 convert negative occurences to 0
H1_NeuMF[H1_NeuMF < 0] = 0
H2_NeuMF[H2_NeuMF < 0] = 0

H1_NeuMF1 <- H1_NeuMF
H2_NeuMF1 <- H2_NeuMF
# remobve the last row
H1_NeuMF1 <- H1_NeuMF[-c(3695), ]
H2_NeuMF1 <- H2_NeuMF[-c(2639), ]
# add the cell names
H1_NeuMF1$X <- H1_Liger$X
H2_NeuMF1$X <- H2_Liger$X
```

```{r}
head(H1_NeuMF1)
head(H2_NeuMF1)
```



# The cell names in one of the datasets does not correspond to the cluster cells.

# Doing the implementation for H2 in liger and neumf

```{r}
library(tidyverse)
library(ggplot2)
```

```{r}
computeError <- function(a, p, n) {
  
  val  = sum( (a-p)^2 )  - sum( (a - n)^2 )
  return(max(val, 0))
}
```



```{r}
computeTripletLoss <- function(embedding, clusters){

  anchor_pos <- clusters %>% filter(x=="B cells")
  hard_neg <- clusters %>% filter(x!="B cells")
  hard_neg_cells <- hard_neg$X
  
  cells <- embedding$X
  
  # this loss variable should be inside
  loss = 0
  
  for (cell in cells){
    # pick a hard positive
    hard_positive <- (sample_n(anchor_pos[anchor_pos$X!=cell, ], 1))$X
    # pick a hard negative
    neg_found = 0
    while (neg_found < 1) {
      hard_negative <- (sample_n(embedding[embedding$X!=cell, ], 1))$X
      if (hard_negative %in% hard_neg_cells) {neg_found = 2}
    }
    
    #now get the vectors associated with these cells
    anchor_vec <- embedding[embedding$X==cell, ]
    pos_vec <- embedding[embedding$X==hard_positive, ]
    neg_vec <- embedding[embedding$X==hard_negative, ]
    
    # compute the square loss
    error <- computeError(anchor_vec[-c(1)], pos_vec[-c(1)], neg_vec[-c(1)]) 
    
    loss = loss + error
    #print(loss)
  }

  #print(paste("loss:", loss))
  return(loss)
}
```


```{r}
iter_loss1 <- rep(0, 100)
iter_loss2 <- rep(1, 100)
index<- c(1:100)
for (k in 1:100){
  iter_loss1[k] <- computeTripletLoss(H2_Liger, original_clusters) 
  iter_loss2[k] <- computeTripletLoss(H2_NeuMF1, original_clusters) 
}
df <- data.frame(iter_loss1)
df$iter_loss2 <- iter_loss2
df$index <- index

ggplot(data = df) + 
  geom_line(aes(x = index, y = iter_loss1, color = "Liger")) + 
  geom_line(aes(x = index, y = iter_loss2, color = "NeuMF")) + 
  scale_color_manual(breaks = c("Liger", "NeuMF"),
                      values = c("green", "blue"))+
  geom_hline( yintercept = mean(iter_loss1), color = "green") + 
  geom_hline(yintercept = mean(iter_loss2), color = "blue")+
  ylab("Loss")
```

```{r}
summary(df)
```

```{r}
clusters <- original_clusters

anchor_pos <- clusters %>% filter(x=="B cells")
  hard_neg <- clusters %>% filter(x!="B cells")
  hard_neg_cells <- hard_neg$X
  
  cells <- stim_dge_t$X1
  
  loss = 0
  
  for (cell in cells[1:50]){
    # pick a hard positive
    hard_positive <- (sample_n(anchor_pos[anchor_pos$X!=cell, ], 1))$X
    # pick a hard negative
    neg_found = 0
    while (neg_found < 1) {
      hard_negative <- (sample_n(stim_dge_t[stim_dge_t$X1!=cell, ], 1))$X1
      if (hard_negative %in% hard_neg_cells) {neg_found = 2}
    }
    
    #now get the vectors associated with these cells
    anchor_vec <- stim_dge_t[stim_dge_t$X1==cell, ]
    pos_vec <- stim_dge_t[stim_dge_t$X1==hard_positive, ]
    neg_vec <- stim_dge_t[stim_dge_t$X1==hard_negative, ]
    
    # compute the square loss
    error <- computeError(anchor_vec[-c(1)], pos_vec[-c(1)], neg_vec[-c(1)]) 
    
    loss = loss + error
    #print(loss)
  }

  #print(paste("loss:", loss))
  return(loss)
```




## F1 - Score

$$F_1 = \frac{TP}{TP + \frac{1}{2}(FP + FN)}$$

### Procedure

(1) Denote the true cell clusters by TC

(2) Extract Liger cell clusters (LC) and neumf cell clusters (NC)

(3) For each cell (c), get it's neighbors in TC (NTC) as well as in LC (NLC) and NC (NNC). 

This procedure does not seem to be making sense.